---
title: "Using Machine Learning for Ocean Chemistry Prediction"
description: Training machine learning models to predict dissolved inorganic carbon in water samples.

editor: visual
author:
  - name: Kiran Favre
    url: https://kiranfavre.github.io
    affiliation: Master of Environmental Data Science
    affiliation-url: https://ucsb-meds.github.io/
date: 03-21-2023
citation: 
  url: https://kiranfavre.github.io/posts/2023-03-21_ml_predictions/ 
categories: [Machine Learning, R, Ocean Chemistry]
format:
  html:
    code-fold: true
    code-summary: "Show code"
---

## Background

For our final project in my machine learning course, we participated in a Kaggle competition to predict the concentration of dissolved inorganic carbon in water samples by using ocean chemistry data. This data comes from the California Cooperative Oceanic Fisheries Investigations (CalCOFI) program.

## Objective

To predict dissolved inorganic carbon, we will be using a Linear Regression Model in R to make these predictions. We will use the CalCOFI data to train our model to make predictions of inorganic dissolved carbon concentrations in different parts of the ocean that aren't included in the training data.

The variables we are using as predictors in our model are:

-   NO2uM - Micromoles of Nitrite per liter of seawater

-   NO3uM - Micromoles of Nitrate per liter of seawater

-   NH3uM - Micromoles of Ammonia per liter of seawater

-   R_TEMP - Reported (Potential) Temperature (degrees Celsius)

-   R_Depth - Reported Depth from pressure (meters)

-   R_Sal - Reported Salinity (from Specific Volume Anomoly, M³ per Kg)

-   R_DYNHT - Reported Dynamic Height (work per unit mass)

-   R_Nuts - Reported Ammonium concentration (micromoles per Liter)

-   R_Oxy_micromol.Kg - Reported Oxygen concentration (micromoles per kilogram)

-   PO4uM - Micromoles of Phosphate per liter of seawater

-   SiO3uM - Micromoles of Silicate per liter of seawater

-   TA1 - Total Alkalinity (micromoles per kilogram solution)

-   Salinity1 - Salinity (Practical Salinity Scale 1978)

-   Temperature_degC - Temperature (degrees Celsius)

## Load and split data

To train machine learning models using a data set, the model must have training data to learn from and test data to compare its predictions to to evaluate model performance. We will then split the training data further into two groups, a validation set and training set. The training set will still be used to train the model, while the validation set will be used to evaluate how well the model performed.

```{r, warning=FALSE, message=FALSE, results='hide', echo = FALSE}
#load libraries 
library(tidyverse)
library(tidymodels)
library(here)
library(janitor)
library(readxl)
```

```{r, warning=FALSE, message=FALSE, results='hide'}
#Reading in data used to train model
training_data <- read_csv(here("posts",
                               "2023-03-21_calcofi_ml",
                               "data",
                               "train.csv")) %>%
  clean_names() %>%
  select(-x13) #remove this since its all NA

#Reading in data that will be used to test model
testing_data <- read_csv(here("posts",
                               "2023-03-21_calcofi_ml",
                               "data",
                               "test.csv")) %>%
  clean_names() %>% 
  mutate(ta1_x = ta1)


#split the training data into training and evaluation sets, stratify by dissolved inorganic carbon concentration
data_split <- initial_split(training_data,
                            strata = dic)

#extract training and test data from the training data
training_set <- training(data_split)
evaluation_set <- testing(data_split)

#take a look at training and testing data 
head(training_data)
head(evaluation_set)
```

## Pre-Processing Data, Creating Recipe, Creating Models, and Creating Workflow

To pre-process the data for our model, we begin by creating a recipe where dissolved inorganic carbon concentration is the predicted value and all the variables mentioned above as the predictors.

```{r}
#set seed for reproducibility
set.seed(711)

#creating a recipe
bottle_recipe <- recipe(dic ~.,
                        data = training_set) %>% 
  step_dummy(all_nominal(),
             -all_outcomes(),
             one_hot = TRUE) %>% 
  step_normalize(all_numeric(),
                 -all_outcomes()) %>% 
  prep()

#creating model specification of linear regression
bottle_model <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")

#bundle recipe and model spec into a workflow
bottle_wf <- workflow() %>% 
  add_recipe(bottle_recipe) %>% 
  add_model(bottle_model)
```

## Fit Model to Training Data and Make Predictions

To understand how well the model we made is performing, we will begin by fitting the model to the training data. This classifies our model as a supervised learner, since we are feeding the model with labeled data to train it. 

We will then use this model to make predictions on the validation data to see how 'well' the model has done, i.e. how accurately is this model predicting values of dissolved inorganic carbon. To evaluate model performance, we will use the Root Mean Square Error (RMSE) metric. The Root Mean Square Error measures the average difference between the actual values and the model's predicted values. For our model, a RMSE of 5.64 indicates that the difference between the actual values and predicted values is quite high. This means our model is not performing well at its task of accurately predicted dissolved inorganic carbon concentrations. 

```{r}
#creating and training a model on the training data
fit_bottle <- bottle_wf %>%
  fit(training_set)

#using the model to make predictions on the validation data   
bottle_results <- fit_bottle %>% 
  predict(evaluation_set) %>%
  bind_cols(evaluation_set) %>% 
  mutate(dic_prediction = .pred_res) %>% 
  relocate(dic,
           .before = id) %>% 
  relocate(dic_prediction,
           .before = id) %>% 
  select(-.pred_res)

#retrieve and evaluate our predictions
bottle_metrics <- bottle_results %>%
  metrics(estimate = dic_prediction,
          truth = dic)

bottle_metrics
```

## Test Model

Here we are predicting the final values for dissolved inorganic carbon and binding them to the ID's of each measuring station. Our professor provided us the actual values for dissolved inorganic carbon on Kaggle to culminate the competition. 
```{r}
## Outputting predictions for our testing data
test_data_predictions <- fit_bottle %>% 
  predict(testing_data) %>%
  bind_cols(testing_data) %>% 
  mutate(DIC = .pred_res) %>% 
  relocate(DIC,
           .before = id) %>% 
  select(id,
         DIC)

test_data_predictions
```
